---
title: "Dose effect multiparasite - Biomass data"
author: "Kit McLean"
date: "08/16/2023"
output: html_document
---

## Biomass GAMs

This document will compute the average mean and standard deviation for the within-host biomass (i.e. spore density * mean spore length within each host) of P. ramosa and M. bicuspidata at time of host death (this is a proxy for transmission) 

```{r , ECHO = FALSE}


library(psych)
library(readxl)
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(data.table)
library(ggpubr)
library(cowplot)

```

Loading data into R:

```{r}

splen <- read.csv("dd_coinf_spore_final.csv")

colnames(splen)[15] <- "finP"
colnames(splen)[19] <- "finM"

#Find biomass mean and SD (Biomass = within-host density at time of host death * average spore length mean or SD)

splen2 <-
  splen %>%
  mutate(m_mean_biomass = finM*m_meanleng, 
         m_sd_biomass = finM*m_sdleng, 
         p_mean_biomass = finP*p_meanleng,
         p_sd_biomass = finP*p_sdleng)

```

Visualizing data: 

A) LENGTH and B) BIOMASS of 1) fungus, and 2) bacterium in A) singly-infected vs B) coinfected hosts: 

```{r}

#One way to subset data by infection status (i.e. assign a character value)

splen2$status <- 0
splen2$status <- as.character(splen2$status)

setDT(splen2)[P > as.numeric("0") & finP > as.numeric("0") & M > as.numeric("0") & finM == as.numeric("0"), status := "PCU"] # Pasteuria only (coinfection unsuccessful)
splen2[M > as.numeric("0") & finM > as.numeric("0") & P > as.numeric("0") & finP == as.numeric("0"), status := "MCU"] #Metsch only (coinfection unsuccessful)
splen2[P == as.numeric("0") & finP == as.numeric("0") | M == as.numeric("0") & finM == as.numeric("0") , status := "US"] #0,0 exposure and a 0,0 infection--there should be NONE of these since we did not include a completely uninfected (0,0) control group 
splen2[P > as.numeric("0") & finP > as.numeric("0") & M == as.numeric("0") & finM == as.numeric("0"), status := "PS"] #Pasteuria infection successful
splen2[M > as.numeric("0") & finM > as.numeric("0") & P == as.numeric("0") & finP == as.numeric("0"), status := "MS"] #Metsch infection successful
splen2[M > as.numeric("0") & finM > as.numeric("0") & P > as.numeric("0") & finP > as.numeric("0"), status := "CS"] #Coinfection successful

#filter out those with missing data
splen2 <- splen2 %>% 
  filter(status != 0)
splen2 <- splen2[!(splen2$status == "MCU" & splen2$p_sdleng > 0),]
splen2 <- splen2[!(splen2$status == "CS" & splen2$m_meanleng == 0),]
splen2 <- splen2[!(splen2$status == "CS" & splen2$p_meanleng == 0),]
splen2 <- splen2[!(splen2$status == "PCU" & splen2$p_meanleng == 0),]

```



Another attempt at visualizing data, first for Metschnikowia:

```{r}

#Metsch mean biomass

ggplot(splen2, aes(x = status, y = m_mean_biomass)) +
  geom_boxplot() + geom_jitter(alpha = 0.4, size = 2)

#Metsch sd biomass

ggplot(splen2, aes(x = status, y = m_sd_biomass)) +
  geom_boxplot() + geom_jitter(alpha = 0.4, size = 2)

#Metsch mean spore length

ggplot(splen2, aes(x = status, y = m_meanleng)) +
  geom_boxplot() + geom_jitter(alpha = 0.4, size = 2)

#Metsch sd spore length

ggplot(splen2, aes(x = status, y = m_sdleng)) +
  geom_boxplot() + geom_jitter(alpha = 0.4, size = 2)
```

Now for Pasteuria...
```{r}
 
#Pasteuria

ggplot(splen2, aes(x = status, y = p_mean_biomass)) +
  geom_boxplot() + geom_jitter(alpha = 0.4, size = 2)

ggplot(splen2, aes(x = status, y = p_sd_biomass)) +
  geom_boxplot() + geom_jitter(alpha = 0.4, size = 2)

ggplot(splen2, aes(x = status, y = p_meanleng)) +
  geom_boxplot() + geom_jitter(alpha = 0.4, size = 2)

ggplot(splen2, aes(x = status, y = p_sdleng)) +
  geom_boxplot() + geom_jitter(alpha = 0.4, size = 2)

#Past has a weirdly broad SD for spore size in coinfected hosts.  Could this be due to shortened host lifespan?

#Check relationship between host lifespan and P size (mean)
Pdat <- subset(splen2, status == "PCU" | status == "PS" | status == "CS") 

ggplot(Pdat, aes(x = lifespan.day, y = p_meanleng, color = status)) +
  geom_point(alpha = 0.4, size = 2)  + geom_smooth(method = "lm")

#... and P size (SD)
ggplot(Pdat, aes(x = lifespan.day, y = p_sdleng, color = status)) +
  geom_point(alpha = 0.4, size = 2)

#Check relationship between host lifespan and M size (mean)
Mdat <- subset(splen2, status == "MCU" | status == "MS" | status == "CS")

ggplot(Mdat, aes(x = lifespan.day, y = m_meanleng, color = status)) +
  geom_point(alpha = 0.4, size = 2)  + geom_smooth(method = "lm")

# ... and M size (SD)
ggplot(Mdat, aes(x = lifespan.day, y = m_sdleng, color = status)) +
  geom_point(alpha = 0.4, size = 2)
```

Now to run a binomial test to see if Coinfection significantly changes the mean or SD of P spore length (as it appears it might when viewing data).

```{r, fig.height=2, fig.width=2.7}

# Convert "coinfection" to 1 and "Past-only infection" to 0
Pdat1 <- Pdat %>%
  mutate(status = recode(status,
                         "CS" = "Coinfected",
                         "PCU" = "Bacterium only",
                         "PS" = "Bacterium only"))

Pdat1$status <- factor(Pdat1$status , levels=c("Bacterium only", "Coinfected"))

#Remove all rows with P biomass == 0 (this is NA data that will inaccurately skew results)
Pdat2 <- Pdat1 %>% 
  filter(p_mean_biomass != 0)

#check for normal distribution of sd data (can I do that?)
Psize_infcat <- ggplot(Pdat2, aes(y = p_meanleng, x = status)) + 
  geom_boxplot(aes(color = status, fill = status), alpha = 0.4, show.legend = FALSE) +
  xlab("Infection class") + 
  ylab("mean spore length (\U003BCm)") +
    scale_color_manual(values = c("#bdc9e1", "#67a9cf")) +
    scale_fill_manual(values = c("#bdc9e1", "#67a9cf")) +
    theme_bw() +
  stat_compare_means(method = "t.test", label.x = 1.22, label.y = 4.5)

Psize_infcat <- 
  ggviolin(Pdat2, x = "status", y = "p_meanleng",
          legend = "none",
         fill = "status", 
         scale = "width",
         palette = c("#bdc9e1", "#67a9cf"),
         add = c("boxplot"),
         add.params = list(fill = "white"),
         xlab = "Infection class",
         ylab = "mean spore length (\U003BCm)",
         font.label = list(size = 11),)  + 
  theme_bw() +
  scale_x_discrete(labels = c("Bacterium\nonly", "Coinfection")) +
  geom_jitter(alpha = 0.4) +
    theme(text = element_text(size=11), legend.position = "none") +
    stat_compare_means(method = "t.test", label.x = 0.8, label.y = 4)

Psize_infcat

t.test(p_meanleng ~ status, Pdat2, var.equal = FALSE)


```
Now for Mestch....
```{r, fig.height=2, fig.width=2.7}
# Convert "coinfection" to 1 and "Metsch-only infection" to 0
Mdat1 <- Mdat %>%
  mutate(status = recode(status,
                         "CS" = "Coinfected",
                         "MCU" = "Fungus only",
                         "MS" = "Fungus only"))
Mdat1$status <- factor(Mdat1$status, levels = c("Fungus only", "Coinfected"))
#Remove all rows with P biomass == 0 (this is NA data that will inaccurately skew results)
Mdat2 <- Mdat1 %>% 
  filter(m_mean_biomass != 0)


#check for normal distribution of sd data (can I do that?)
Msize_infcat <- 
  ggviolin(Mdat2, x = "status", y = "m_meanleng",
          legend = "none",
         fill = "status", 
         scale = "width",
         palette = c("#02818a", "#67a9cf"),
         add = c("boxplot"),
         add.params = list(fill = "white"),
         xlab = "Infection class",
         ylab = "mean spore length (\U003BCm)",
         font.label = list(size = 11),)  + 
  theme_bw() +
  scale_x_discrete(labels = c("Fungus\nonly", "Coinfection")) +
  geom_jitter(alpha = 0.4) +
    theme(text = element_text(size=11), legend.position = "none") +
    stat_compare_means(method = "t.test", label.x = 0.8, label.y = 41)

Msize_infcat


t.test(m_meanleng ~ status, data = Mdat2, var.equal = FALSE)

M_Coinf_var_test <- Mdat2 %>%
  kruskal.test(m_meanleng ~ status)


```

Length & biomass of 1) fungus vs. A) bacterium (within coinfected hosts; be sure to plot paired data & account for pairing in stats):
```{r}

succ0b <- 
  splen2 %>%
  filter(P == as.numeric("0") & finP == as.numeric("0")|
        M == as.numeric("0") & finM == as.numeric("0"))
           
succPb <- 
  splen2 %>%
  filter(P > as.numeric("0") & finP > as.numeric("0"))

succMb <-  
  splen2 %>%
  filter(M > as.numeric("0") & finM > as.numeric("0"))

succCoinfb <-
  splen2 %>%
  filter(M > as.numeric("0") & finM > as.numeric("0") & P > as.numeric("0") & finP > as.numeric("0"))

succb <- do.call("rbind", list(succ0b, succPb, succMb, succCoinfb))

ggplot(succCoinfb, aes(x = p_mean_biomass, y = m_mean_biomass)) +
  geom_point() 

ggplot(succCoinfb, aes(x = p_meanleng, y = m_meanleng)) +
  geom_point() 

#Pearson correlation coefficient for paired samples
cor.test(succCoinfb$p_meanleng, succCoinfb$m_meanleng, method = "pearson")
```

Length of 1) fungus and 2) bacterium vs. A) fungus dose and B) bacterium dose (within coinfected hosts; be sure to plot paired data & account for pairing in stats):
```{r}

ggplot(succCoinfb, aes(x = M, y = m_meanleng)) +
  geom_point() 

ggplot(succCoinfb, aes(x = P, y = p_meanleng)) +
  geom_point() 

ggplot(succCoinfb, aes(x = finM, y = m_meanleng)) +
  geom_point() 

ggplot(succCoinfb, aes(x = finP, y = p_meanleng)) +
  geom_point() 


#Pearson correlation coefficient for paired samples
Pearson_M_Mleng <- cor.test(succCoinfb$M, succCoinfb$m_meanleng, method = "pearson")
Pearson_M_Mleng

Pearson_P_Pleng <- cor.test(succCoinfb$P, succCoinfb$p_meanleng, method = "pearson")
Pearson_P_Pleng


```

Length of 1) fungus and 2) bacterium vs. A) fungus dose and B) bacterium dose (within any host)
```{r}

Mdat2$M <- as.factor(Mdat2$M)

Msize_Mdose <- Mdat2 %>%
  ggplot(aes(x = M, y = m_meanleng)) +
  geom_boxplot() + 
  theme_bw() +
  xlab("Fungal dose (spores/mL)") + 
  ylab("mean spore length (\U003BCm)") 

Pdat2$P <- as.factor(Pdat2$P)

Psize_Pdose <- Pdat2 %>%
  ggplot(aes(x = P, y = p_meanleng)) +
  geom_boxplot()  + 
  theme_bw() +
  xlab("Bacterial dose (spores/mL)") + 
  ylab("mean spore length (\U003BCm)")


#Pearson correlation coefficient for paired samples
Pearson_M_Mleng <- cor.test(Mdat$M, Mdat$m_meanleng, method = "pearson")
Pearson_M_Mleng

Pearson_P_Pleng <- cor.test(Pdat$P, Pdat$p_meanleng, method = "pearson")
Pearson_P_Pleng

#Or we can treat dose as a categorical variable
Pdat3 <- Pdat2
Pdat3$P <- as.character(Pdat3$P)
summary(aov(p_meanleng ~ P, data = Pdat3))

Mdat3 <- Mdat2
Mdat3$M <- as.character(Mdat3$M)
summary(aov(m_meanleng ~ M, data = Mdat3))

```

Biomass of 1) fungus and 2) bacterium vs.host lifespan in A) single infections and C) coinfections:
```{r}

# #Metsch 
# plot <- ggplot(df[which(df$prop>0),], aes(X5employff, prop)) +
#   geom_bar(aes(fill=X5employff, stat="identity")) +
#   facet_grid( ~ X5employf, scales="free_x") +
#   theme_bw()

#Trying to use "which" so I don't have to subset data
Msize_lifespan<-
  ggplot(Mdat1[which(Mdat1$M>0 & Mdat1$finM>0),], aes(x = lifespan.day, y = m_meanleng, color = status)) +
  geom_point(alpha = 0.4, size = 2)  + geom_smooth(method = "lm", aes(fill = status)) +
  xlab("Host lifespan") + 
  ylab("mean spore length (\U003BCm)") +
    scale_color_manual(values = c("#02818a", "#67a9cf")) +
    scale_fill_manual(values = c("#02818a", "#67a9cf")) +
    theme_bw() +
    theme(legend.position = c(.75, .25))

ggplot(splen2[which(splen2$M>0 & splen2$finM>0),], aes(x = lifespan.day, y = finM)) +
  geom_point()


#Pasteuria 
Psize_lifespan <-
  ggplot(Pdat1[which(Pdat1$P>0 & Pdat1$finP>0),], aes(x = lifespan.day, y = p_meanleng, color = status)) +
  geom_point(alpha = 0.4, size = 2)  + 
  geom_smooth(method = "lm", aes(fill = status)) +
  xlab("Host lifespan") + 
  ylab("mean spore length (\U003BCm)") +
    scale_color_manual(values = c("#bdc9e1", "#67a9cf")) +
    scale_fill_manual(values = c("#bdc9e1", "#67a9cf")) +
    theme_bw() +
    theme(legend.position = c(.75, .25))

Psize_lifespan

ggplot(splen2[which(splen2$P>0 & splen2$finP>0),], aes(x = lifespan.day, y = finP)) +
  geom_point()


## Stats looking for sig relationship between spore size and lifespan
Pearson_life_Mleng <- cor.test(Mdat$M, Mdat$lifespan.day, method = "pearson")
Pearson_life_Mleng

Pearson_life_Pleng <- cor.test(Pdat$P, Pdat$lifespan.day, method = "pearson")
Pearson_life_Pleng


```
Using cowplot to create a 6 panel figure
```{r, fig.height= 5, fig.width= 4}
 plot_grid(Psize_Pdose, Msize_Mdose, Psize_lifespan, Msize_lifespan, Psize_infcat, Msize_infcat, labels = "AUTO", align = "h", ncol = 2)
 
 
```
