---
title: "Analysis of data on spore dose & multiparasite exposures/infections"
author: "Kit McLean & Meghan A. Duffy"
date: "July 2021-June 2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T,
                      results = "hide", 
                      warning = FALSE, 
                      message = FALSE)
```

# Coinfection response surface study 



```{r, results = 'hide'}
library(ggplot2)
library(gridExtra)
library(dplyr)
library(mgcv)
library(lubridate)
library(psych)
library(grid)
library(animation)
library(itsadug)
library(readxl)
library(tidyr)
library(stringr)
library(tinytex)
library(scales)
library(ggpubr)
library(gratia)
library(patternplot)
library('gratia')
library(rstatix)
library(coin)
library(emmeans)

# Analyses to Run
# A.Probability of Infection
# B.Host lifespan 
# C.Within-host spore load

#Loading data into R...
rawdat <- read.csv("Coinfection_Response_Surface_main.csv")

```


```{r message=FALSE, include=FALSE}
 
## Data Clean-up

#assign simpler column names to main variable of interest
colnames(rawdat)[1:2] <- c("P", "M")
colnames(rawdat)[10] <- "finP"
colnames(rawdat)[14] <- "finM"

rawdat$P <- as.numeric(rawdat$P)
rawdat$M <- as.numeric(rawdat$M)

#Drop weird NAs at end of data frame (must be an import error; these are not present in .csv file itself)
rawdat2 <-
  rawdat %>%
  drop_na(P,M) %>%
  subset(select= -c(6:9, 11:13,15:17))

#Convert dates to a workable format
Exp.date <- mdy(rawdat2$Exposure.date)
Mort.date <- mdy(rawdat2$Mortality.date)
as.numeric(Exp.date)
as.numeric(Mort.date)

#Now I want to find the length of time it took each animal to die in days.
timetodeath <- Exp.date %--% Mort.date
timetodeath <- as.period(timetodeath)
timetodeath.sec <- period_to_seconds(timetodeath)

#animals were infected at 6 days old, therefore I'm adding 518400 seconds (6 days) to the timetodeath to get total lifespan
lifespan.sec <- timetodeath.sec + 518400
lifespan.day.frac <- lifespan.sec/86400
lifespan.day <- ceiling(lifespan.day.frac)

dat1 <- cbind(rawdat2, lifespan.day)

#Replace extremely low spore counts with 0, as I assume those were cross-contaminations
dat1$finM <- as.numeric(dat1$finM)
dat1$finP <- as.numeric(dat1$finP)

dat1$finM[dat1$finM < 6668] <- 0
dat1$finP[dat1$finP < 6668] <- 0

dat2 <- dat1 %>%
  filter(Exposure.date != "NA") %>%
  mutate(finM = replace(finM, finM == "#DIV/0!", 0))

#Change "NA" mortality dates to 0 (these are uninfected animals)
dat2$Mortality.date[is.na(dat2$Mortality.date)] <- 0

#Convert extremely low spore yields to spore yield = 0 (because )

#Create columns for binary data on infection status (1 = TRUE, 0 = FALSE)
dat2$P_yes = (dat2$`finP` > 0)
dat2$M_yes = (dat2$`finM`> 0)
dat2$Co_yes =  (dat2$`finP` > 0) & (dat2$`finM`> 0)
dat2$Uninf = (dat2$`Mortality.date` == 0 | dat2$`finP` == 0 & dat2$`finM`== 0)


dat2$P_yes <- as.numeric(dat2$P_yes)
dat2$M_yes <- as.numeric(dat2$M_yes)
dat2$Co_yes <- as.numeric(dat2$Co_yes)
dat2$Uninf <- as.numeric(dat2$Uninf)

dat3 <- dat2 %>% filter(Mortality.date != "missing")

#Split off uninfected animals from the rest of the data set so it's easier to clean up
#the infected data set

uninfecteddat <- dat3 %>% filter(Mortality.date == 0)

#Infected only dat, minus all missing values
infecteddat <- dat3 %>% filter(Mortality.date != 0,
                               finP != "missing",
                               finP != "Missing",
                               finM != "missing",
                               finM != "Missing")

dat4 <- rbind(uninfecteddat, infecteddat)

#Make sure spore counts are numeric
dat4$finM <- as.numeric(dat4$finM)
dat4$finP <- as.numeric(dat4$finP)

#Create new column for animals that were diagnosed as infected but had 0 spores when processed
dat4$ZeroSpore = (dat4$`Mortality.date` !=0) & (dat4$`finP` == 0) & (dat4$`finM` == 0)
dat4$ZeroSpore <- as.numeric(dat4$ZeroSpore)

dat4[is.na(dat4)] <- 0

#Removing the "oops" coinfection
dat4 <- dat4[!(dat4$M == 0 & dat4$finM != 0),]

#Convert spore densities into per-host densities (by dividing by 10)
dat5 <- 
  dat4 %>%
  mutate(
    finM = finM/10,
    finP = finP/10
  )

#Save as its own dataset
final.fulllifespandat <- dat5

#Rename and filter to exclude hosts that were diagnosed as infected but had 0 spores when processed
final <- dat5 %>% filter(ZeroSpore == 0)

#transform the data into log version so it is more usable for GAMs
transfinal <- 
  final %>% 
  mutate(logaP = log(P+1),
         logaM = log(M+1))

#The gam needs synthetic training data to operate properly at spore doses = 0 (so the model shows a probability of infection equal to (very near) 0 when there are 0 parasites present)
#The gam needs synthetic training data to operate properly at spore doses = 0

 P <- rep(c(0), each = 10000)
 M <- rep(c(0), each = 10000)
 finP <- rep(c(0), each = 10000)
 finM <- rep(c(0), each = 10000)
 logaP <- rep(c(0), each = 10000)
 logaM <- rep(c(0), each = 10000)
 P_yes <- rep(c(0), each = 10000)
 M_yes <- rep(c(0), each = 10000)
 Co_yes <- rep(c(0), each = 10000)
 Uninf <- rep(c(0), each = 10000)
 Infect <- rep(c(0), each = 10000)

dfzeros <- data.frame(P,M,finP,finM,logaP,logaM,P_yes,M_yes,Co_yes,Uninf,Infect)

transzeros <- full_join(transfinal, dfzeros)

```

## A. PROBABILITY OF INFECTION  
### Data Exploration & GAMs

### Dose Past and/or Metsch dose determine probability of infection with Past(only), Metsch(only), or Coinfection(both)?

### Answer: 
### Pr(Past infection) determined by P & M ; model: Pinf_gam
### Pr(Metsch infection) determined by M ; model: Minf_gam
### Pr(Coinfection) determined by P & M ; model: Coinf_gam
 
```{r}

Pnull_dat <- 
  transzeros %>%
  filter(M == "0")

#Probability of Past infections (single + coinfections) without interaction terms
Pinf_gam <- gam(P_yes ~ s(logaP, k = 7) + s(logaM, k = 7), method = "REML", data=transzeros, family = binomial)

plot(Pinf_gam,se = TRUE) #se stands for standard error Bands

#Using a GAM approach with interaction terms
Pinf_gam_inter <- gam(P_yes ~ s(logaP, k = 7) + s(logaM, k = 7) + te(logaP, logaM, k = c(7,7)), method = "REML", data=transzeros, family = binomial)

#Comparing the two models
summary(Pinf_gam)
summary(Pinf_gam_inter)

#The interaction term is not significant, so we plot the simpler model.

vis.gam(Pinf_gam, main= "Probability of any bacterial infection", plot.type = "contour", type= "response",
       color = "topo", contour.col = "black", lwd = 2, xlab = expression(paste("ln(bacterial dose (spores / mL))")), ylab = expression(paste("ln(fungal dose (spores / mL))")))

##############################
```

```{r}

#probability of Metsch infections (single + coinfections) 
Minf_gam <- gam(M_yes ~ s(logaP, k = 7) + s(logaM, k = 7), method = "REML", data=transzeros, family = binomial)

plot.gam(Minf_gam)

##Using a GAM approach with interaction terms
Minf_gam_inter <- gam(M_yes ~ s(logaP, k = 7) + s(logaM, k = 7) + te(logaP, logaM, k = c(7,7)), method = "REML", data=transzeros, family = binomial)

#Comparing the two models
summary(Minf_gam)
summary(Minf_gam_inter)

#The interaction term is not significant, so we plot the simpler model.
vis.gam(Minf_gam, view = NULL, main= "vis.gam: Probability of any fungal infection", plot.type = "contour", type= "response", color = "topo", contour.col = "black", lwd = 2, xlab = expression(paste("ln(bacterial dose (spores / mL))")), ylab = expression(paste("ln(fungal dose (spores / mL))")))

```

```{r}

#probability of coinfection is determined by both P dose and M dose
Coinf_gam <- gam(Co_yes ~ s(logaP, k=7) + s(logaM, k=7), method = "REML", data=transzeros)

##Using a GAM approach with interaction terms
Coinf_gam_inter <- gam(Co_yes ~ s(logaP, k = 7) + s(logaM, k = 7) + te(logaP, logaM, k = c(7,7)), method = "REML", data=transzeros, family = binomial)

#Comparing the two models
summary(Coinf_gam)
summary(Coinf_gam_inter)

#The interaction term is not significant, so we plot the simpler model.
vis.gam(Coinf_gam, main= "Probability of coinfection", plot.type = "contour", type= "response",
       color = "topo", contour.col = "black", lwd = 2, xlab = expression(paste("ln(bacterial dose (spores / mL))")), ylab = expression(paste("ln(fungal dose (spores / mL))")))


```

```{r}

#Independent or interaction effects on prevalence of coinfection? (Using real data)

propfinal <- transfinal %>%
  group_by(P,M) %>%
  summarise(
    propP = mean(P_yes == "1"),
    propM = mean(M_yes == "1"),
    propCo = mean(Co_yes == "1")
  ) %>%
  mutate(logaP = log(P+1),
         logaM = log(M+1))


InterEff <- glm(((propCo/(propP * propM)-1)) ~ logaP + logaM + logaP*logaM, data = propfinal)
summary(InterEff)

#Drop ns interaction
InterEff0 <- glm(((propCo/(propP * propM)-1)) ~ logaP + logaM, data = propfinal)
summary(InterEff0)

#Drop ns main effects
InterEff1 <- glm(((propCo/(propP * propM))) ~ 1, data = propfinal)
summary(InterEff1)

#So proportion of coinfected is about half of what you would predict. 
#I.e. it looks like there is some interference between the two parasites at the point of infection  

``` 


## B.0.	HOST LIFESPAN (incomplete measure of virulence) 
### Data exploration 
#### with log-transformed dose data

```{r, fig.width=6, fig.height= 6}

#Assign infection categories (IU = diagnosed as infected but had 0 spores during spore counts, P = Pasteuria-only infections, M = Metsch-only infections, C = Coinfection with Past and Metsch) and remove all uninfected animals, which don't have lifespan data (Mortality.date !=0)

translifedat2 <-
  final.fulllifespandat %>% mutate(infect.cat = case_when(
    finP == 0 & finM == 0 & Mortality.date !=0 ~ "IU" ,
    finP > 0 & finM == 0 ~ "P"  ,
    finP == 0 & finM > 0 ~ "M"  ,
    finP > 0 & finM > 0 ~ "C"
  )
) %>%
  mutate(logaP = log(P+1),
         logaM = log(M+1)) %>%
  filter(Mortality.date != 0)

lifedat2.nozs <-
  translifedat2 %>% filter(infect.cat != "IU")

life_infectcat_viol <-
  ggviolin(lifedat2.nozs, x = "infect.cat", y = "lifespan.day", 
         fill = "infect.cat",
         scale = "width",
         palette = c("#02818a", "#bdc9e1", "#67a9cf"),
         #add = c("boxplot"),
         add.params = list(fill = "white"), 
         xlab = " ",
         ylab = "Lifespan (days)", 
         show.legend = FALSE
         ) + 
  theme_bw() +
  theme(text = element_text(size=11), axis.text = element_text(size = 11), legend.position = "none") +
  scale_x_discrete(labels = c("Fungus", "Bacterium", "Coinfected")) + 
   theme(text = element_text(size=11), axis.text = element_text(size = 11)) +
  geom_jitter(alpha = 0.7, show.legend = FALSE) +
    ylim(10,80) 



```


## B.INFECTED HOST LIFESPAN (incomplete measure of virulence)
### Data analysis

### B.1. What is the relationships between host lifespan and parasite dose?

#### Answer: Lifespan is determined mostly by infection class (IU, P, M, or C), but also by log Metsch dose
```{r}
###  Host lifespan ~ Pathogen dose  ###

lifedoseinfcat_nouninf_log <- glm(lifespan.day ~ infect.cat + logaM + logaP + logaM*logaP, data = lifedat2.nozs, family = poisson())

summary(lifedoseinfcat_nouninf_log)

# Interaction isn't significant, so removing that
lifedoseinfcat_nouninf_log_noint <- glm(lifespan.day ~ infect.cat + logaM + logaP, data = lifedat2.nozs, family = poisson())

summary(lifedoseinfcat_nouninf_log_noint)
drop1(lifedoseinfcat_nouninf_log_noint,test="Chisq")


### Follow up analysis: what if we drop logaP and look at an interaction between infect.cat & logaM?

followupinteractionmodel <- glm(lifespan.day ~ infect.cat * logaM, data = lifedat2.nozs, family = poisson())

summary(followupinteractionmodel)
drop1(followupinteractionmodel,test="Chisq")
```

```{r, plot to explore effect of metsch dose on lifespan}

MetschDoseLifespanPlot <- 
  ggplot(lifedat2.nozs, aes(logaM, lifespan.day, shape=infect.cat, colour=infect.cat, fill=infect.cat)) +
  geom_smooth(method="lm") +
  geom_point(size=3) +
  ylim(10,80) +
  ylab("Lifespan (days)") +
  xlab("Log(fungal spore dose)") +
#  ggtitle("Lifespan ~ bacterial spore yield")  + 
   theme_bw() +
  scale_fill_manual(values = c("#67a9cf", "#02818a", "#bdc9e1"), labels=c('coinf','fungus','bact')) +
  scale_color_manual(values=c("#67a9cf","#02818a", "#bdc9e1"), labels=c('coinf','fungus','bact')) +
  scale_shape_discrete(labels=c('coinf','fungus','bact')) +
    theme(legend.position = c(0.75, 0.85), legend.key.size = unit(0.25, 'cm'))


MetschDoseLifespanPlot


lifedat2.nozs.justP <- lifedat2.nozs %>%
  filter(infect.cat == "P")

lifedat2.nozs.justM <- lifedat2.nozs %>%
  filter(infect.cat == "M")

lifedat2.nozs.justC <- lifedat2.nozs %>%
  filter(infect.cat == "C")

cor.test(lifedat2.nozs.justP$logaM, lifedat2.nozs.justP$lifespan.day)

cor.test(lifedat2.nozs.justM$logaM, lifedat2.nozs.justM$lifespan.day)

cor.test(lifedat2.nozs.justC$logaM, lifedat2.nozs.justC$lifespan.day)


```

```{r, making a combined lifespan figure}
combinedlifespanplot <- ggarrange(life_infectcat_viol, MetschDoseLifespanPlot, 
          labels = c("A", "B"),
          align = c("h"),
          ncol = 2, nrow = 1, common.legend = FALSE)

combinedlifespanplot


ggsave("combinedlifespanplot.png", combinedlifespanplot, width = 6, height = 3.5, dpi = 300)
```

B.2. Is lifespan significantly different between infection classes?

#### Answer: Yes.
```{r}
###  Host lifespan ~ Infection class ###

#NOTE TO KDM: I changed this to emmeans, which gives the ratio, which I think is handy
lifeinfcat_contrasts <- emmeans(lifedoseinfcat_nouninf_log_noint, specs = pairwise ~ infect.cat, type = "response")
lifeinfcat_contrasts

#Mean number of days coinfected hosts lived: 
Co_lifespan_mean <- 
  translifedat2 %>%
  filter(infect.cat == "C")

mean(Co_lifespan_mean$lifespan.day)

#Mean number of days Past-only infected hosts lived: 
Past_lifespan_mean <- 
  translifedat2 %>%
  filter(infect.cat == "P")

mean(Past_lifespan_mean$lifespan.day)

#Mean number of days Metsch-only infected hosts lived: 
Metsch_lifespan_mean <- 
  translifedat2 %>%
  filter(infect.cat == "M")

mean(Metsch_lifespan_mean$lifespan.day)

```


#C: SPORE YIELD

## Violin plots of final spore yield
```{r}
altdat <- translifedat2

altwinP <- 
  altdat %>%
  filter(finP > 0)

altwinM <- 
  altdat %>%
  filter(finM > 0)

myield_viol <-
  ggviolin(altwinM, x = "infect.cat", y = "finM", 
         fill = "infect.cat",
         scale = "width",
         palette = c("#02818a", "#67a9cf"),
         #add = c("boxplot"),
         add.params = list(fill = "white"), 
         xlab = " ",
         ylab = "Fungal spore yield (spores per host)", 
         show.legend = FALSE
         ) + 
  theme_bw() +
  theme(text = element_text(size=11), axis.text = element_text(size = 11), legend.position = "none") +
  scale_x_discrete(labels = c("Fungus", "Coinfected")) + 
   theme(text = element_text(size=11), axis.text = element_text(size = 11)) +
  geom_jitter(alpha = 0.7, show.legend = FALSE) 

myield_viol


pyield_viol <-
  ggviolin(altwinP, x = "infect.cat", y = "finP", 
         fill = "infect.cat",
         scale = "width",
         palette = c( "#bdc9e1","#67a9cf"),
         #add = c("boxplot"),
         add.params = list(fill = "white"), 
         xlab = " ",
         ylab = "Bacterial spore yield (spores per host)", 
         show.legend = FALSE
         ) + 
  theme_bw() +
  theme(text = element_text(size=11), axis.text = element_text(size = 11), legend.position = "none") +
  scale_x_discrete(labels = c("Bacterium","Coinfected")) + 
   theme(text = element_text(size=11), axis.text = element_text(size = 11)) +
  geom_jitter(alpha = 0.7, show.legend = FALSE) 

pyield_viol
```

## Does infection class (single v coinf) impact the final spore yield? 

```{r}
#Subset so all animals are infected with Past (single or coinf)
transwinhostP <- transfinal %>% filter(P_yes == 1)
#Subset so all animals are infected with Metsch (single or coinf)
transwinhostM <- transfinal %>% filter(M_yes == 1)


#Model effects of Coinfection, P dose, and M dose on finM (ie M yield)
Mdens.parasiteID <- glm(finM ~ Co_yes + logaM + logaP, data = transwinhostM)

#Model effects of Coinfection, M dose, and P dose on finP (ie P yield)
Pdens.parasiteID <- glm(finP ~ Co_yes + logaM + logaP, data = transwinhostP) 

summary(Mdens.parasiteID)
summary(Pdens.parasiteID)

#Finding the percent decrease in within-host density of Past due to coinfection with Metsch
WinP_mean_Ponly <- 
  transwinhostP %>%
  filter(Co_yes == 0)

mean(WinP_mean_Ponly$finP)

WinP_mean_Coonly <- 
  transwinhostP %>%
  filter(Co_yes == 1)

mean(WinP_mean_Coonly$finP)

(mean(WinP_mean_Ponly$finP) - mean(WinP_mean_Coonly$finP)) / mean(WinP_mean_Ponly$finP)

#Finding the percent increase in within-host density of Metsch due to coinfection with Past
WinM_mean_Monly <- 
  transwinhostM %>%
  filter(Co_yes == 0)

mean(WinM_mean_Monly$finM)

WinM_mean_Coonly <- 
  transwinhostM %>%
  filter(Co_yes == 1)

mean(WinM_mean_Coonly$finM)

(mean(WinM_mean_Coonly$finM) - mean(WinM_mean_Monly$finM)) / mean(WinM_mean_Monly$finM)

```

### Plots of lifespan vs. spore yield
```{r}
LifespanMetschYieldPlot <- 
  ggplot(altwinM, aes(lifespan.day, finM, shape=infect.cat, colour=infect.cat, fill=infect.cat)) +
  geom_smooth(method="lm") +
  geom_point(size=3) +
  ylab("Fungal spore yield (spores per host)") +
  xlab("Lifespan (days)") +
   theme_bw() +
  scale_fill_manual(values = c("#67a9cf", "#02818a"), labels=c('coinf','fungus')) +
  scale_color_manual(values=c("#67a9cf","#02818a"), labels=c('coinf','fungus')) +
  scale_shape_discrete(labels=c('coinf','fungus')) +
    theme(legend.position = c(0.75, 0.85), legend.key.size = unit(0.25, 'cm'))

LifespanMetschYieldPlot

LifespanPastYieldPlot <- 
  ggplot(altwinP, aes(lifespan.day, finP, shape=infect.cat, colour=infect.cat, fill=infect.cat)) +
  geom_smooth(method="lm") +
  geom_point(size=3) +
  ylab("Bacterial spore yield (spores per host)") +
  xlab("Lifespan (days)") +
   theme_bw() +
  scale_fill_manual(values = c("#67a9cf","#bdc9e1"), labels=c('coinf','bact')) +
  scale_color_manual(values=c("#67a9cf","#bdc9e1"), labels=c('coinf','bact')) +
  scale_shape_manual(values = c(19,15), labels=c('coinf','bact')) +
    theme(legend.position = c(0.75, 0.85), legend.key.size = unit(0.25, 'cm'))

LifespanPastYieldPlot
```

```{r, making a combined spore yield figure}
combinedsporeyieldplot <- ggarrange(pyield_viol, myield_viol, LifespanPastYieldPlot,LifespanMetschYieldPlot,
          labels = c("A", "B","C","D"),
          align = c("h"),
          ncol = 2, nrow = 2, common.legend = FALSE)

combinedsporeyieldplot


ggsave("combinedsporeyieldplot.png", combinedsporeyieldplot, width = 7, height = 7, dpi = 300)
```

```{r, lifespan spore yield correlations}

#Subsetting for all animals infected with Past only
transpastonly <- transfinal %>% filter(finP > 0 & finM == 0) %>% 
  mutate(logaP = log(P+1),
         logaM = log(M+1))
#Subsetting for all animals infected with Metsch only
transmetschonly <- transfinal %>% filter(finP == 0 & finM > 0) %>% 
  mutate(logaP = log(P+1),
         logaM = log(M+1))
#Subsetting for all animals coinfected
transcoinf <- transfinal %>% filter(finP > 0 & finM > 0) %>% 
  mutate(logaP = log(P+1),
         logaM = log(M+1))

cor.test(transpastonly$lifespan.day,transpastonly$finP)

cor.test(transmetschonly$lifespan.day,transmetschonly$finM)

cor.test(transcoinf$lifespan.day,transcoinf$finP)

cor.test(transcoinf$lifespan.day,transcoinf$finM)
```


## C.3. Is there a correlation between M spore yield and P spore yield?   

```{r}

PYieldMYieldPlot <- 
  ggplot(transcoinf, aes(finP, finM)) +
  geom_smooth(method="lm", color = c("#67a9cf"), fill = c("#67a9cf")) +
  geom_point(size=3, color = c("#67a9cf"), fill = c("#67a9cf")) +
  xlab("Bacterial spore yield (spores per host)") +
    ylab("Fungal spore yield (spores per host)") +
   theme_bw() +
  scale_fill_manual(values = c("#67a9cf")) +
  scale_color_manual(values=c("#67a9cf")) +
  scale_shape_manual(values = c(19)) 

PYieldMYieldPlot

ggsave("PYieldMYieldPlot.png", PYieldMYieldPlot, width = 4, height = 3.5, dpi = 300)

cor.test(transcoinf$finP, transcoinf$finM)

```


